---
title: "Intro Stats for PHP - Practical 4"
author: Daniel J Carter
output: html_notebook
---

This R Notebook corresponds to Practical 4 in Stata for Intro Stats for PHP. R Notebooks are like Stata do files but more flexible: they act as both text documents and code files so there is no need to comment everything in the same way! 

In Stata, most functions you might require for data analysis are already built into the program. R is open source and therefore users of R contribute their own useful functions to accomplish set tasks. These are contained in packages. Once you install a package it sits on your computer, so you only need install it once. You must however tell R that you need that suite of functions by loading its library. The below two commands do this. We need to do it for three packages for this practical.

You can run commands in R by placing your cursor anywhere in the line and pressing Ctrl-Enter (no need to highlight like Stata).

```{r}
#--- Install the tidyverse package and load library
install.packages("tidyverse")
library(tidyverse)

#--- Install the foreign package and load library
install.packages("foreign")
library(foreign)

#--- Install the psych package and load library
install.packages("psych")
library(psych)
```

Let's read in the data. In R, rather than specify the working directory, you can simply place the dataset in the same folder as the R Notebook you are working in. Then the below command will run.

The arrow is an assignment operator. It tells R to assign the name on the left of the arrow to the object on the right. So the below reads in the data and then stores the dataset as an object named 'heights'. You should see it appear in the top right corner.

```{r}
#--- Read in the data
heights <- read.dta("./heights.dta")

#--- Look at the data
View(heights)
head(heights) # this is the same as 'list height in 1/6'
```

Let's look at some summaries. Note the syntax. describe() is the verb, and it is describing the noun heights$height. Recall that heights is the dataset. The dollar sign tells R that we want the column 'height' from that dataset.

```{r}
#--- Describe data (from psych package)
describe(heights$height)
```

In this chunk of code, we calculate the 95% reference range. The qnorm() function gives the z value for a given probability (e.g. 0.05 in two tails maps to about 1.96)

```{r}
#--- Input values
mean <- mean(heights$height)
sd <- sd(heights$height)
n <- 20

#--- Exact calculation of reference range
mean + qnorm(.975) * sd/sqrt(n)
mean - qnorm(.975) * sd/sqrt(n)

#--- Approximate calculation
84.25 + 1.96 * 3.16/sqrt(20)
84.25 - 1.96 * 3.16/sqrt(20)

```


Here we calculate the CI and conduct a t-test at the same time. R does not have a built in function for a z-test as it is so rarely used, so I provide a t-test below. There are z tests available in other packages (good exercise to install for yourself!)

```{r}
#--- Student's t-test
t.test(heights$height, mu = 86.5)
```

We would interpret this test as providing strong evidence against the null  hypothesis that the average height of Jamaican boys with sick cell differs from the UK mean height of 86.5.

---

Now we turn to the skinfold tests. In R, we can have multiple datasets in memory.

```{r}
#--- Read in the data
skinfold <- read.dta("./PRAC4_Q2.dta")
```

We use the tidyverse to create a new variable. The %>% is what's known as a pipe, like in Super Mario. It takes what's on the left and puts it into the right. So here, we do not need to continually type the dataset name: R knows which dataset we are working with.

```{r}
#--- Generate new variable
skinfold <- skinfold %>% 
  mutate(diff = skinmeasure_observer_a - skinmeasure_observer_b)

#--- Look at new variable
describe(skinfold$diff)

#--- Conduct the t-test
t.test(skinfold$diff)

```

The interpretation of this test is that there is strong evidence against the null hypothesis that there is no difference in the observer values. Observer A recorded higher measurements on average.

--- 

Moving to Part 3: 

```{r}
#--- Read in the data
titres <- read.dta("./PRAC4_Q3.dta")

#--- Get the approximate reference range
describe(titres$titre)

16.2 + 1.96*(26.96)
16.2 - 1.96*(26.96)

#--- Generate nice plot
titres %>% 
  ggplot(aes(x = titre)) + # tell ggplot to put titre on x axis
  geom_histogram() + # tell ggplot you want a histogram
  theme_bw() # use a black and white theme

#--- Generate the log data (note in R log() = ln; log10() = log)
titres <- titres %>% 
  mutate(ln.titre = log(titre))

#--- Describe and generate plot
describe(titres$ln.titre)

titres %>% 
  ggplot(aes(x = ln.titre)) + 
  geom_histogram() + 
  theme_bw()

#--- Get the geometric mean and CIs for log transformed values
ttest <- t.test(titres$ln.titre)

#--- Extract the upper and lower bounds from the test itself
lowerCI <- exp(ttest$conf.int[1])
upperCI <- exp(ttest$conf.int[2])
geom_mean <- exp(mean(titres$ln.titre))

#--- Print the results (NB: c() stands for concatenate)
print(c(lowerCI, upperCI, geom_mean))

```



