---
title: "2: Review of logistic regression"
subtitle: "R 4 ASME"
author: Author – Andrea Mazzella & Daniel J Carter
---

-------------------------------------------------------------------------------

## Contents

This is a summary of the four SME topics on logistic regression.

-------------------------------------------------------------------------------

## Data management

```{r message=FALSE, warning=FALSE}
# Load packages
library(haven)
library(magrittr)
library(epiDisplay)
library(epitools)
library(pubh)
library(rstatix)
library(tidyverse)

# Limit significant digits to 2, remove scientific notation
options(digits = 2, scipen = 9)
```

```{r}
# Data import
mwanza <- read_dta("mwanza17.dta") %>% as_factor()

# Data tidying
# Recode missing values
mwanza %<>% mutate(across(everything(), ~ na_if(.x, 9)))

# Create new variables, relevel and label
mwanza %<>%
  mutate(age2 =  as.factor(
    case_when(
      age1 == "1" | age1 == "2"  ~ "15-24",
      age1 == "3" | age1 == "4" ~ "25-34",
      age1 == "5" | age1 == "6" ~ "35+"))) %>% 
  mutate(ed2 =  as.factor(
    case_when(ed == 1 ~ 0,
              ed != 1 ~ 1)))
```


# 1. Planning your model

_Without coding_, write a logistic regression model to investigate the association between:
- HIV status (outcome)
- lifetime number of sexual partners (`npa`) as a 4-level factor

Then build on this model by including schooling (`ed2`) as a binary variable.


# 2. Logistic regression

## 2a. Tabulation

> Obtain a frequency table of `npa`.

```{r}
mwanza %$% table(npa, useNA = "ifany")
```

What is the most common number of lifetime sexual partners?

> Cross-tabulate number of lifetime sexual partners with HIV status.

```{r}
mwanza %$% tabpct(npa, case, percent = "col", graph = F)
```



## 2b. Unadjusted logistic regression

> Fit a logistic model to estimate the magnitude of association between `npa` (as a factor) and HIV status.

```{r}
glm(case ~ npa, family = "binomial", data = mwanza) %>% 
  logistic.display()

```

There is very strong evidence against the null hypothesis of no association between HIV status and number of sexual partners (LRT p < 0.001).



## 2c. Change baseline group

By default, the baseline level of comparison will be the smallest value. You might want to use the most prevalent level of `npa` as a baseline, in order to calculate OR relative to that level.

In order to do this, you need to relevel the factor - here is the most explicit way of doing so:

```{r}
# Relevel the factor
mwanza$npa <- factor(mwanza$npa,
                     levels = c("2", "1", "3", "4"))

# Logistic regression (unchanged)
glm(case ~ npa, family = "binomial", data = mwanza) %>% 
  logistic.display()

# Relevel the factor back, if you want
# mwanza$npa <- factor(mwanza$npa,
#                      levels = c("1", "2", "3", "4"))
```


## 2d. Logistic model with confounding

> Now also include `age1` treated as a factor in your model (keeping 2 as the baseline level).

```{r}
glm(case ~ npa + age1, family = "binomial", data = mwanza) %>% 
  logistic.display()

```

What is your conclusion?

# 3. More on confounding + intro to interaction

## 3a. School

> Check if the risk of HIV associated with `npa` and `age1` is confounded by attending school (`ed2`).

```{r}
glm(case ~ npa + age1 + ed2, family = "binomial", data = mwanza) %>% 
  logistic.display()
```


## 3b. 
In R formulae, the `*` (equivalent to `##` in Stata) is used to specify including both an interaction AND individual risk factors. We use lrtest in the epiDisplay package to compare models. What do you conclude?

```{r}
# Model with interaction
logit_inter <- glm(case ~ npa * ed2 + age1, family = "binomial", data = mwanza)

# Model without interaction
logit_without <- glm(case ~ npa + ed2 + age1, family = "binomial", data = mwanza)

# Likelihood ratio test
epiDisplay::lrtest(logit_without, logit_inter)

```

# 4. Interaction with more than 2 levels

> Try fitting a model including an interaction between `npa` and `age1` and have a look at the results.

```{r}
glm(case ~ npa * age1, family = "binomial", data = mwanza) %>% 
  logistic.display()
```


Cross-tabulate `npa` and `age1`. What's the problem and how can we solve it?

```{r}
mwanza %$% table(age1, npa, case, useNA = "ifany")
```


## 4b. Solving the issue

In order to fix the issue of data sparsity, we can combine levels 3 and 4 of `npa`. 

```{r}
# Create a new variable, relevel and label
mwanza %<>%
  mutate(partners =  factor(
    case_when(npa == "1" ~ "<=1",
              npa == "2" ~ "2-4",
              npa == "3" | npa == "4" ~ ">=5"),
    levels = c("2-4", "<=1", ">=5")
  ))

# Check it worked well
mwanza %$% table(npa, partners, useNA = "ifany")
```

We can then use this new variable, `partners`, to create a model for interaction and compare it to a model without interaction with a LRT.

```{r}
# Model with interaction
logit_inter3 <- glm(case ~ partners * age1, family = "binomial", data = mwanza)
epiDisplay::logistic.display(logit_inter3)

# Model without interaction
logit_without3 <- glm(case ~ partners + age1, family = "binomial", data = mwanza)

# Likelihood ratio test
epiDisplay::lrtest(logit_inter3, logit_without3)

```



# 5. 28/04 -- To come :)



-------------------------------------------------------------------------------

