---
title: "8: Cox regression in cohort studies"
subtitle: "R 4 ASME"
authors: Authors – Lakmal Mudalige, Andrea Mazzella, Daniel J Carter
output: html_notebook
---

-------------------------------------------------------------------------------

## Contents

* Cox regression models
  * simple
  * adjusting for follow-up and age
  * adjusting for other covariates
* Comparing Poisson and Cox
 
NB: the time-to-event (Kaplan-Meier) approach is not used here, but it was explored with the same dataset in SME.
 
-------------------------------------------------------------------------------

## 0. Packages and options

```{r message=FALSE, warning=FALSE}
# Load packages
library(haven)
library(magrittr)
library(survival)
library(fmsb)
library(jstable) # new package! May have to install
library(epiDisplay)
library(lubridate)
library(tidyverse)

# Limit significant digits to 3, reduce scientific notation
options(digits = 3, scipen = 9)
```

-------------------------------------------------------------------------------


# Trinidad

## 1. Data manipulation

Read in dataset trinmlsh.dta. This contains data from a cohort study on cardiovascular risk factors and mortality among ~300 men from Trinidad - see the dataset descriptions for more information.

```{r}
# Import the dataset
trin <- read_dta("trinmlsh.dta")

# Preview
trin
```

Categorical variables need to be factorised and labelled, and later in the practical, we will need to regroup some smoking levels, so this is included in data management.

```{r data management, include=FALSE}
trin %<>%
  mutate(
    ethgp = factor(ethgp,
                      levels = c(1:5),
                      labels = c("African", "Indian", "European", "mixed", "Chin")),
    alc = factor(alc,
                      levels = c(0:3),
                      labels = c("none", "1-4/wk", "5-14/wk", ">=15/wk")),
    smokenum = factor(smokenum,
                      levels = c(0:5),
                      labels = c("non-smok", "ex-smok", "1-9/d", "10-19/d", "20-29/d", ">=30/d")),
    chdstart = factor(chdstart,
                      levels = c(0, 1),
                      labels = c("no", "yes")))

# Regroup smoking
trin %$% table(smokenum, useNA = "ifany")
trin$smok3 <- as.factor(ifelse(trin$smokenum == "non-smok" , "non-smok",
                        ifelse(trin$smokenum == "ex-smok", "ex-smok","smoker"))) %>%
              fct_relevel("non-smok", "ex-smok", "smoker")
trin %$% table(smokenum, smok3, useNA = "ifany")

```

Now create a survival object to assess all-cause mortality, coded as `death`. Note that Surv() can take time data in two different formats: either a combination of data of entry and data of exit (like in the previous praticas), or as a time difference. In this case, `years` codes this time difference, so we'll use it.

```{r}
# Survival object
trin_surv <- trin %$% Surv(time = years, event = death)
```

-------------------------------------------------------------------------------


## 2. Cox regression

We can now examine the smoking-specific mortality rates (per 1,000 person-years). Let's first use the classical technique and then let's use Cox regression.

```{r}
# Calculate rates & rate ratios
pyears(trin_surv ~ smok3, data = trin, scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)

rateratio(18, 30, 462, 984, conf.level=0.95)
rateratio(40, 30, 789, 984, conf.level=0.95)

```

The package "survival" contains `coxph()`, the function for Cox regression. Exactly like `pyears()`, `coxph()` takes as first argument a formula with the survival object on the left and the exposure on the right.

```{r}
# Create Cox model
coxph(trin_surv ~ smok3, data = trin)
coxph(trin_surv ~ smok3, data = trin) %>% jstable::cox2.display()
```

The mortality rate ratio appears to increase for each level of smoking; there is some weak evidence against the null hypothesis of equal hazard ratios (LRT p = 0.07).

-------------------------------------------------------------------------------


## 3. Cox with a numerical exposure

So, let's analyse the same association but with smoking coded as quantitative.

```{r}
# Cox with numeric exposure
coxph(trin_surv ~ as.numeric(smok3), data = trin)
coxph(trin_surv ~ as.numeric(smok3), data = trin) %>% jstable::cox2.display()
```
Now there is some evidence against the null hypothesis of no linear trend: from one level of smoking to the next, the hazard increases by about 1.3 times.

-------------------------------------------------------------------------------


## 4. Timescale set as current age

Cox regression always automatically adjusts for a time variable: the way we made our survival object earlier, the timescale was set as time since entry. But what if adjusting for current age gave us different results?

This could happen if there are differences in age between the smoking levels. Let's check this.

```{r}
#--- Quantify
trin %>% group_by(smok3) %>%
         drop_na() %>% 
         summarise("N" = n(),
                   "1st quartile" = quantile(ageent, 0.25),
                   "median" = median(ageent),
                   "3rd quartile" = quantile(ageent, 0.75))


#-- Visualise
trin %>% drop_na() %>% 
  ggplot(aes(x = smok3, y = ageent)) + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_jitter(aes(x = reorder(smok3, desc(ageent)), y = ageent, colour = as.factor(death)), width=0.15) +
  labs(title = "Age at entry by smoking status",
       y = "Age at entry (years)", 
       x = "", 
       colour = "CHD death") + 
  theme_bw() + 
  scale_colour_brewer(type = "qual", palette = 2)
```

The age of entry seems similar in all three levels, so the results shouldn't be very different with a different timescale, but just for completeness, we check anyway.

To set the timescale as current age, we need to put the date of birth in the "origin" argument and convert dates to numbers of years (since 01/01/1970, the base date in R). 

```{r}
# Survival object set for current age
trin_surv_age <- trin %$% Surv(as.numeric(timein) / 365.25, 
                               as.numeric(timeout) / 365.25,
                               death,
                               origin = as.numeric(timebth) / 365.25)

# Cox using the timeage as the time 
coxph(trin_surv_age ~ smok3, data = trin)
coxph(trin_surv_age ~ smok3, data = trin) %>% jstable::cox2.display()
```

-------------------------------------------------------------------------------


# Primary Biliary Cholangitis (PBC)

This dataset contains information from an RCT comparing the immunosuppressant azathioprine with placebo.

* Outcome: `death`
* Treatment: `treat`
* Time:
 * `time` (follow-up in years)
 * `age` in years
* Other variables:
 * `logb0` (log serum bilirubin concentration)
 
Given these time variables, how would you set up your survival object?

```{r include=FALSE}
# Read in the pbc1bas.dta dataset 
pbc <- read_dta("pbc1bas.dta")

glimpse(pbc)

# Data management
pbc %<>% mutate(death = d,
               treat = factor(treat, levels = c(1, 2), labels = c("placebo", "azath")),
               cenc0 = factor(cenc0, levels = c(0, 1), labels = c("no", "yes")),
               cir0 = factor(cir0, levels = c(0, 1), labels = c("no", "yes")),
               gh0 = factor(gh0, levels = c(0, 1), labels = c("no", "yes")),
               asc0 = factor(asc0, levels = c(0, 1), labels = c("no", "yes"))) %>% 
         select(-d)

glimpse(pbc)

summary(pbc)
```

-------------------------------------------------------------------------------


## 5. Comparing Poisson and Cox

Assess the relationship between treatment and mortality adjusting for baseline bilirubin, first with Poisson, and then with Cox regression.

```{r}
# Poisson model
glm(death ~ offset(log(time)) + treat + logb0, family = poisson(), data = pbc) %>% idr.display()

# Cox model
cox_pbc <- survival::coxph(Surv(time, death) ~ treat + logb0, data = pbc, model = TRUE)
jstable::cox2.display(cox_pbc)

```

Poisson: HR 0.73 (0.49, 1.10), LRT p = 0.13
Cox:     HR 0.65 (0.43, 0.99), LRT p = 0.04

Unlike Poisson, Cox regression also adjusts for time in study, and its results provide good evidence for an effect of azathioprine. A great exercise for the mathematically inclined of you reading this is to demonstrate that a Cox proportional hazards model is exactly equivalent to a Poisson model under the assumption of constant baseline hazard.

-------------------------------------------------------------------------------


## 6. Cox'ing the Poisson

In order to make the Poisson model more similar to Cox, we need to account for time in study. We do this by performing a Lexis expansion - this time, instead of doing the Lexis expansion by hand as in the previous practical, we will use the functions from the 'Epi' package, and then add the newly-created "period" categorical variable as another covariate.

```{r splitLexis}
library(Epi)

# Create a Lexis object
pbc$time.in <- 0
Lexis.pbc <-
  pbc %>% Lexis(
    entry = list(per = time.in),
    exit = list(per = time),
    exit.status = death,
    data = .
  )

# Lexis split by period
Lexis.pbc.per <-
  splitLexis(Lexis.pbc, breaks = c(0, 2, 4, 6), time.scale = "per")

# Create a variable with the period
Lexis.pbc.per$per <- timeBand(Lexis.pbc.per, "per", type = "factor")

# Fit a Poisson model to the split Lexis object
PBC.m2 <-
  glm(
    lex.Xst ~ offset(log(lex.dur)) + per + as.factor(treat) + logb0,
    family = poisson(),
    data = Lexis.pbc.per
  )

exp(PBC.m2$coefficients)
confint(PBC.m2) %>% exp()
```

-------------------------------------------------------------------------------


# No such thing as too much Whitehall

Read in whitehal.dta.

```{r include=FALSE}
whitehall <- read_stata("whitehal.dta")

# Factorise job grade
whitehall$grade <- factor(whitehall$grade,
                          levels = c(1, 2),
                          labels = c("higher", "lower"))

whitehall %<>% mutate(py = time_length(interval(as.Date(timein), as.Date(timeout)), "days"))

glimpse(whitehall)
```

Build a Cox model to estimate the effect of job grade (`grade`) on cardiac mortality (`chd`), first with the follow-up scale. Afterwards, assess whether age is a confounder with both Cox and Poisson techniques. The time variables are `timein`, `timeout` and `timebth`.

```{r}
# Survival object in follow-up scale
surv_wh <- whitehall %$% Surv(time = as.numeric(timein) / 365.25, 
                                 time2 = as.numeric(timeout) / 365.25, 
                                 event = chd)

# Cox
coxph(surv_wh ~ grade, data = whitehall) %>% cox2.display()
```

Now, let's assess if age is a confounder by changing the timescale.

```{r}
# Age as timescale
coxph(Surv(timein, timeout, chd, origin = timebth) ~ grade, data= whitehall) %>% cox2.display()
```

This HR is much lower than the previous one, indicating that age is likely a confounder for the effect of grade. Note that this also makes sense from a causal perspective: age is almost certainly a common cause of both mortality and civil service grade.

The same analysis with Poisson requires us to split the data into smaller chunks of time and estimate within those chunks.

```{r}
# Poisson with no Lexis expansion
glm(chd ~ offset(log(follow.up)) + grade, family = poisson(), data = whitehall) %>% idr.display()

# Do the Lexis expansion
wh.lexis <- whitehall %>% 
  Lexis(entry = list(per=timein), exit = list(per=timeout, age = timeout - timebth), exit.status = chd, data = .) %>% 
  splitLexis(breaks = c(0, 40, seq(50,80,5)), time.scale="age") 

wh.lexis$currage <- timeBand(wh.lexis, "age", type = "factor")

# Poisson regression with Lexis expansion
glm(lex.Xst ~ offset(log(lex.dur)) + currage + grade, family = poisson(), data = wh.lexis) %>% idr.display()
```

Now we see very similar results -- accounting for age in this way makes Poisson regression closely resemble a Cox regression.

-------------------------------------------------------------------------------
