---
title: "1: Classical analysis of categorical variables"
subtitle: "R 4 ASME"
author: Authors – Andrea Mazzella & Daniel J Carter
---

Welcome to R for ASME! If you are brand new to R, we recommend working through the introductory material first: Visualising Data & Summarising Data at a minimum, followed by Tidying Data and Modelling Data. These will give you a brief introduction to the R language and some common epidemiological tasks. Many basic R concepts are NOT introduced in these practicals, unlike in the earlier practicals for STEPH and SME and their knowledge is assumed - the first half of this practical also acts as revision of some core R concepts. There is a forum available on Moodle for any questions on R-related content. 

-------------------------------------------------------------------------------

## Contents - Practical 1

* Basic data management
  * import
  * explore
  * clean
  * tidy
  * export

* Classical statistical analysis of categorical variables
  * chi-squared test
  * test for trend
  * OR
  * adjusted OR
  * OR with logistic regression

-------------------------------------------------------------------------------

## 0. Load packages

```{r message=FALSE, warning=FALSE}
# Load packages
library(haven)
library(magrittr)
library(epitools)
library(epiDisplay)
library(pubh)
library(rstatix)
library(tidyverse)

# Limit significant digits to 2, reduce scientific notation
options(digits = 2, scipen = 9)
```


# Basic data management

## 1. Change the default directory

If you use an .Rmd notebook like this one, you don't need to change directory – R assumes the default directory is the directory where the .Rmd is kept. So, you need to put your datasets in the same folder as this notebook. We recommend using R Projects to keep all your datafiles in one place (and for easy Github integration). Chapter 8 of R 4 Data Science gives an overview of how to use R Projects: https://r4ds.had.co.nz/workflow-projects.html


## 2. Import and explore a .dta dataset

The function `read_dta()` is from package {haven}. Note that for these R practicals, there are slightly different versions of the dataset required to appropriately red in categorical variables. Please ensure you are using the versions of the dataset with the '17' affix.

```{r}
mwanza <- read_dta("mwanza17.dta") %>% as_factor()
```

Unfortunately, Stata help files cannot be accessed in R. Key points regarding this dataset:

- Case-control study (not matched)
- Cases: women aged >=15 years living with HIV
- Controls: random sample of HIV-negative women
- Cases and controls were interviewed about potential risk factors for HIV
- All variables are categorical and there are no value labels

To look at the variables and their types:

```{r}
# Variable names
names(mwanza)

# Preview of the dataset
mwanza

# Visualise the whole dataset
View(mwanza)
```

```{r include=FALSE}
# Variable names, types, and first values
glimpse(mwanza)
```

## 3. Function syntax and getting help

The general syntax in R is:

`function(argument1, argument2)`

Arguments are always separated by a comma.

R can use multiple datasets at the same time, which means that you always need to specify the dataset on which you undertake most operations. With the _tidyverse_, this can be written as:

`dataset %>% function()`

```{r}
# These two are equivalent
names(mwanza)
mwanza %>% names()
```

To use a variable, you still need to specify which dataset it is from:
`function(dataset$variable1)`

With package {magrittr}, this can be written as:
`dataset %$% function(variable1)`

For example, to tabulate education by HIV status with column percents, you can use the `tabpct()` function in _epiDisplay_.

```{r}
# These two are equivalent
tabpct(mwanza$ed, mwanza$case, graph = F)
mwanza %$% tabpct(ed, case, graph = F)
```

Unlike Stata, you can't abbreviate function or variable names. But if you start typing a variable in a dataset, RStudio will guess what it is and you can select it by pressing Tab and then Enter.

You can filter rows and select columns with functions from package {dplyr}, part of the {tidyverse}.

For example, to do the same tabulation but only in those aged less than 30, you would type:

```{r}
mwanza %>%
  filter(age1 <= 3) %$%
  tabpct(ed, case, graph = F)
```

To do the same tabluation across multiple variables (here: age1, education, religion), we can use the across() function within a summarise() function. The syntax here is slightly complicated but the key part is that the tilde (~) tells R to expect a function (tabpct) -- and then the ".x" stands for "apply that function to the variables in x" -- where the variables in x come from the first argument to across.

```{r}
mwanza %>% 
  summarise(across(c(age1, ed, rel), 
                   ~ tabpct(.x, case, percent = "col", graph = F)))
```

To get help about a function or a package, you put a question mark and then the name of that function/package:

```{r}
? View
```

There is also a lot of help online - StackExchange is a very helpful reseource!


## 4. Saving the results

Whenever you save the .Rmd notebook, a record will be created in the same folder – its file format will depend on what's written in the "output:" field at the top of the .Rmd and the default is html.


## 5. Creating and recoding variables

The variable `age1` is coded in groups:
1: 15-19, 2: 20-24, 3: 25-29, 4: 30-34, 5: 35-44, 6: 45-54

The variable `ed` represents year of education, coded as 1: none, 2: 1-3 years, 3: 4-6 years, 4: 7+ years

The variable `ed2` is binary; 0: none, 1: 1+ years. The variable `npa` represents lifetime number of sexual partners, coded as 1: 0-1, 2: 2-4, 3: 5-9, 4: 10+, 9: missing.

To create a new variable (or change an existing one), you use the function `mutate()`. You can combine this with `as.factor()` and `case_when()` to relevel a categorical variable and give labels to the values in one single bit of code. This function says "if the condition on the left of the tilde (~) is met, then assign the label on the right.

You can then cross-tabulate with `table(..., useNA = "ifany")` to ensure this went well.

```{r}
# Create a new variable, relevel and label
mwanza %<>%
  mutate(age2 =  as.factor(case_when(age1 <= 2 ~ "15-24",
                                     age1 == 3 | age1 == 4 ~ "25-34",
                                     age1 == 5 | age1 == 6 ~ "35+")))

# Create a new variable, relevel and label
mwanza %<>%
  mutate(ed2 =  as.factor(case_when(ed == 1 ~ 0,
                                     ed != 1 ~ 1)))

# Ensure it went ok
mwanza %$% table(age1, age2, useNA = "ifany")
```

Missing values are recorded as NA in R. To tell R to treat the value of 9 as missing for variable `npa`, we use (using the %<>% pipe to combine <- and %>%):

```{r}
mwanza %<>% mutate(npa = na_if(npa, 9))

# Check it worked ok
mwanza %$% table(npa, useNA = "ifany")
```

You can use `mutate()` to recode more than one variable at once, and `map()` to apply the tabulation function to more than one variable in one command.

```{r}
mwanza %<>% mutate(msta = na_if(msta, 9),
                   eth = na_if(eth, 9),
                   rel = na_if(rel, 9))

mwanza %>%
  select(msta, eth, rel) %>%
  map(table, useNA = "ifany")
```


## 6. Saving the current dataset

Don't overwrite the existing dataset, just make a copy.

```{r}
# Stata
mwanza %>% write_dta("mwanza2.dta")
```

-------------------------------------------------------------------------------


# Classical statistical analysis of categorical variables


## 7. Chi-squared testing

In EpiDisplay, you can get a chi-square test with the 'cc' command

What is the association between education and HIV status?

```{r}
# Chi-square
mwanza %$% cc(ed, case, graph = F)

```


## 8. Test for trend

Is there a linear association?

`prop_trend_test()` is from package _rstatix_. Note that this just provides the test for trend - you should investigate the 

```{r}
# Test for trend
mwanza %$% table(case, ed) %>%
  prop_trend_test()
```


## 9. Odds ratios

We can calculate the stratum-specific odds ratios for each level of a variable using the table() command and the oddsratio() command from _epitools_.

```{r}
# Stratified ORs
mwanza %$% table(ed, case) %>% oddsratio() %$% measure
```

It looks like more educated people have higher odds of having HIV than less educated people.


## 10. Adjusted odds ratios

Let's now check if this association still remains after accounting for age - note that am specifying epiDisplay:: here as the packaged pubh also contains a function called mhor with slightly different syntax.

```{r}
mwanza %$% epiDisplay::mhor(case, ed2, age2, graph = F)
```


## 11. Odds ratios using logistic regression

You can also use logistic regression without covariates to get the OR across all levels.
`logistic.display()` makes the output of logistic regression easier to read, and more similar to Stata. 

```{r}
# Logistic regression
glm(case ~ ed,
    family = "binomial",
    data = mwanza) %>%
  epiDisplay::logistic.display()
```

You can also add a covariate by adding it after `+`.
Note that the output provides not only the adjusted OR, but also the crude OR (without the covariate) 

```{r}
# Logistic regression with covariate
glm(case ~ ed + age2,
    family = "binomial",
    data = mwanza) %>%
  epiDisplay::logistic.display()
```

-------------------------------------------------------------------------------
